---
title: "Class 12"
author: "Emma Bell (A19247017)"
format: pdf
toc: true
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble.

Here we read the CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
mxl$Genotype..forward.strand.
```
```{r}
table(mxl$Genotype..forward.strand.)
```

So for Q.5 9, people have G/G. But that is not the proportion

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl)
```

SO! it is 0.140625


# Section 4. Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)

```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```
These are the samples sizes for Q.13

```{r}
summary(expr)
```
```{r}
library(dplyr)

expr %>%
  group_by(geno) %>%
  summarize(median_exp = median(exp, na.rm = TRUE))
```

These are the median expression levels for Q13

```{r}
library(ggplot2)
```

Let's make a boxplot


```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```



Having a G/G correlates with having a reduced expression.
Therefore the SNP does have an effect on the expression of ORMDL3






